
---
title: "Moderation Analysis: Gaming Hours, Unemotional Traits, and Aggression"
author: "Generated by Databot"
date: "2026-02-11"
format: 
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    theme: default
execute:
  warning: false
  message: false
---

::: {.callout-note}
This report was generated using AI under general human direction. At the time of generation, the contents have not been comprehensively reviewed by a human analyst.

<!--
To indicate human review: Delete the line above about contents not being reviewed, and replace this comment with:
The contents have been reviewed and validated by [Your Name], [Your Role] on [Date].
-->
:::

## Introduction

This report presents a moderation analysis examining whether **unemotional traits** moderate the relationship between **hours spent gaming** and **aggression**. Moderation analysis helps us understand when and for whom the effect of a predictor variable (gaming hours) on an outcome variable (aggression) is stronger or weaker based on the level of a third variable (unemotional traits).

The research question is: Does the relationship between gaming hours and aggression depend on an individual's level of unemotional traits?

## Data Loading and Preparation

First, we load the necessary packages and the data:

```{r}
library(ggplot2)
library(knitr)

# Load the data
aggression <- read.csv("2026-02-11_VU_AI_show_and_share/agression.csv")

# Display first few rows
head(aggression)
```

The dataset contains `{r} nrow(aggression)` observations with three variables:

- **hours_spent_gaming**: Number of hours spent gaming
- **unemotional_traits**: Score on unemotional traits measure
- **aggression**: Aggression score

### Centering Predictors

For moderation analysis, we center the predictor and moderator variables. Centering helps with interpretation by making the main effects represent the effect at the mean level of the other variable.

```{r}
# Center the predictors
aggression$hours_centered <- scale(aggression$hours_spent_gaming, scale = FALSE)
aggression$unemotional_centered <- scale(aggression$unemotional_traits, scale = FALSE)

# Create the interaction term
aggression$interaction <- aggression$hours_centered * aggression$unemotional_centered
```

## Moderation Analysis

We fit a regression model with the centered predictors and their interaction:

```{r}
# Fit the moderation model
moderation_model <- lm(aggression ~ hours_centered + unemotional_centered + interaction, 
                       data = aggression)

# Display the results
summary(moderation_model)
```

### Results Table

```{r}
# Create a comprehensive results table
results_table <- data.frame(
  Predictor = c("Intercept", 
                "Hours Spent Gaming (centered)", 
                "Unemotional Traits (centered)", 
                "Interaction"),
  Coefficient = sprintf("%.3f", coef(moderation_model)),
  SE = sprintf("%.3f", summary(moderation_model)$coefficients[, 2]),
  t_value = sprintf("%.3f", summary(moderation_model)$coefficients[, 3]),
  p_value = sprintf("%.4f", summary(moderation_model)$coefficients[, 4]),
  Significance = c("***", "***", "*", "***")
)

kable(results_table, align = c("l", "r", "r", "r", "r", "c"),
      caption = "Moderation Analysis: Predicting Aggression")
```

**Note:** \* p < .05, \*\* p < .01, \*\*\* p < .001

### Key Findings

The moderation analysis reveals:

- **Interaction effect**: The interaction term is highly significant (β = `{r} sprintf("%.3f", coef(moderation_model)[4])`, p < .001), indicating that unemotional traits significantly moderate the relationship between gaming hours and aggression.

- **Main effect of gaming hours**: β = `{r} sprintf("%.3f", coef(moderation_model)[2])`, p < .001

- **Main effect of unemotional traits**: β = `{r} sprintf("%.3f", coef(moderation_model)[3])`, p = .050 (marginally significant)

## Model Fit Statistics

```{r}
# Model fit statistics table
model_fit <- data.frame(
  Statistic = c("R²", "Adjusted R²", "F-statistic", "df", "p-value", "Residual SE"),
  Value = c(
    sprintf("%.4f", summary(moderation_model)$r.squared),
    sprintf("%.4f", summary(moderation_model)$adj.r.squared),
    sprintf("%.2f", summary(moderation_model)$fstatistic[1]),
    sprintf("%d, %d", summary(moderation_model)$fstatistic[2], 
            summary(moderation_model)$fstatistic[3]),
    "< 0.0001",
    sprintf("%.3f", summary(moderation_model)$sigma)
  )
)

kable(model_fit, align = c("l", "r"),
      caption = "Model Fit Statistics")
```

The model explains `{r} sprintf("%.1f%%", summary(moderation_model)$r.squared * 100)` of the variance in aggression scores and is highly significant overall.

## Simple Slopes Analysis

To understand the moderation effect, we examine the simple slopes - the effect of gaming hours on aggression at different levels of unemotional traits:

```{r}
# Calculate simple slopes at different levels of the moderator
sd_unemotional <- sd(aggression$unemotional_traits)
mean_unemotional <- mean(aggression$unemotional_traits)
coefs <- coef(moderation_model)

# Simple slope = b1 + b3*moderator_value_centered
simple_slopes <- data.frame(
  Moderator_Level = c("Low (-1 SD)", "Mean", "High (+1 SD)"),
  Unemotional_Value = c(mean_unemotional - sd_unemotional,
                        mean_unemotional,
                        mean_unemotional + sd_unemotional),
  Simple_Slope = c(
    coefs[2] + coefs[4] * (-sd_unemotional),
    coefs[2] + coefs[4] * 0,
    coefs[2] + coefs[4] * sd_unemotional
  )
)

# Create interpretive table
simple_slopes_table <- data.frame(
  `Unemotional Traits Level` = simple_slopes$Moderator_Level,
  `Moderator Value` = sprintf("%.2f", simple_slopes$Unemotional_Value),
  `Simple Slope` = sprintf("%.3f", simple_slopes$Simple_Slope),
  `Interpretation` = c(
    "Negative: More gaming → Less aggression",
    "Positive: More gaming → More aggression",
    "Strong Positive: More gaming → Much more aggression"
  )
)

kable(simple_slopes_table, align = c("l", "r", "r", "l"),
      caption = "Simple Slopes: Effect of Gaming Hours at Different Levels of Unemotional Traits")
```

### Interpretation of Simple Slopes

The simple slopes reveal a striking pattern:

- **At low levels of unemotional traits** (-1 SD): The slope is **negative** (`{r} sprintf("%.3f", simple_slopes$Simple_Slope[1])`), meaning that more gaming hours are associated with *lower* aggression.

- **At mean levels of unemotional traits**: The slope is **positive** (`{r} sprintf("%.3f", simple_slopes$Simple_Slope[2])`), indicating a positive relationship between gaming and aggression.

- **At high levels of unemotional traits** (+1 SD): The slope is **strongly positive** (`{r} sprintf("%.3f", simple_slopes$Simple_Slope[3])`), showing that more gaming hours are associated with substantially higher aggression.

## Visualization

The interaction effect is best understood visually:

```{r}
#| fig-width: 10
#| fig-height: 7

# Calculate means for centering
mean_hours <- mean(aggression$hours_spent_gaming)
mean_unemotional <- mean(aggression$unemotional_traits)
sd_unemotional <- sd(aggression$unemotional_traits)

# Define moderator levels
moderator_levels <- c(
  "Low (-1 SD)" = mean_unemotional - sd_unemotional,
  "Mean" = mean_unemotional,
  "High (+1 SD)" = mean_unemotional + sd_unemotional
)

# Create hours range
hours_range <- seq(min(aggression$hours_spent_gaming), 
                   max(aggression$hours_spent_gaming), 
                   length.out = 100)

# Calculate predictions for each moderator level
plot_data <- data.frame()
for(i in 1:length(moderator_levels)) {
  hours_centered <- hours_range - mean_hours
  unemotional_centered <- moderator_levels[i] - mean_unemotional
  interaction <- hours_centered * unemotional_centered
  
  predicted <- coefs[1] + 
               coefs[2] * hours_centered + 
               coefs[3] * unemotional_centered + 
               coefs[4] * interaction
  
  temp <- data.frame(
    hours_spent_gaming = hours_range,
    predicted_aggression = predicted,
    moderator_level = names(moderator_levels)[i]
  )
  plot_data <- rbind(plot_data, temp)
}

# Create the plot
ggplot(plot_data, aes(x = hours_spent_gaming, y = predicted_aggression, 
                      color = moderator_level, linetype = moderator_level)) +
  geom_line(linewidth = 1.2) +
  labs(title = "Moderation Effect: Unemotional Traits Moderate Gaming-Aggression Relationship",
       x = "Hours Spent Gaming",
       y = "Predicted Aggression",
       color = "Unemotional Traits",
       linetype = "Unemotional Traits") +
  theme_minimal() +
  theme(legend.position = "bottom",
        legend.title = element_text(size = 11),
        plot.title = element_text(size = 12, face = "bold"))
```

The plot clearly shows:

1. **Diverging lines**: The three lines fan out, which is the visual signature of an interaction effect
2. **Crossover interaction**: The lines actually cross, indicating a reversal of the relationship
3. **Direction reversal**: At low unemotional traits (green dashed line), more gaming predicts less aggression, while at high unemotional traits (red solid line), more gaming predicts much more aggression

## Interpretation and Conclusions

### Summary of Key Findings

This moderation analysis reveals that **the relationship between gaming hours and aggression is critically dependent on an individual's level of unemotional traits**:

1. **Significant Moderation Effect**: The interaction between gaming hours and unemotional traits is highly significant (p < .001), explaining `{r} sprintf("%.1f%%", summary(moderation_model)$r.squared * 100)` of the variance in aggression.

2. **Context-Dependent Effect**: The effect of gaming on aggression is not uniform:
   - For individuals **low in unemotional traits**, gaming is associated with *lower* aggression
   - For individuals **high in unemotional traits**, gaming is associated with *substantially higher* aggression

3. **Practical Implications**: This finding suggests that:
   - Gaming itself may not be universally problematic for aggression
   - Individual characteristics (specifically unemotional traits) determine whether gaming is associated with increased or decreased aggression
   - Interventions targeting gaming-related aggression should consider individual differences in emotional processing

### Methodological Notes

- Predictors were centered to aid interpretation
- The interaction term captures how the effect of one variable changes as a function of another
- Simple slopes analysis decomposes the interaction at meaningful levels of the moderator (-1 SD, mean, +1 SD)

### Limitations

- This analysis assumes linear relationships
- Causal interpretation requires experimental or longitudinal data
- Additional covariates may affect the relationship

---

**Report generated:** `{r} format(Sys.Date(), "%B %d, %Y")`
